Functional Pearl: Short and Mechanized Logical Relation for Dependent Type Theories

### 2024-12-13 paper/paper.tex
Minor edits through Section 8

@@

The $\beta\eta$-normalization proof from Section~\ref{sec:betaeta} comprises
1549 lines of non-blank, non-comment lines of code. We choose not to include
it in the chart, because of slight differences in lemma dependencies for
untyped reduction and normal forms that make the comparison less
informative. However, when compared to the $\beta$-normalization extension,
the $\beta\eta$ extension has the same line count in the definition of the
logical relation and the semantic soundness proof.

@@

We use the Autosubst 2 framework to produce Coq syntax files based on a de
Bruijn representation of variable binding and capture-avoiding substitution.
In addition to these generated definitions, Autosubst 2 provides a powerful
tactic \texttt{asimpl} that can be used to prove the equivalence of two terms
constructed using the primitive operators provided by the framework. This
tactic simplifies the reasoning about substitution as many
substitution-related properties about syntax are immediately discharged by
\texttt{asimpl}.

For other automation tasks that are not specific to binding, we use
the powerful \texttt{sauto} tactic provided by CoqHammer to write
short and declarative proofs. For example, here is a one-line proof of
the triangle property about parallel reduction, from which the diamond
property (Lemma~\ref{lemma:pardiamond}) follows as a corollary.
The triangle property states
that if $[[a => b]]$, then $[[b]]\Rightarrow [[a]]^*$, where $[[a]]^*$
is the Takahashi translation~\citep{takahashi-parallel-reduction}
which roughly corresponds to simultaneous reduction of the redexes in
$[[a]]$, excluding the new redexes that appear as a result of
reduction.
\begin{minted}[escapeinside=||,mathescape=true]{coq}
Lemma Par_triangle a : forall b, (a |$\Rightarrow$| b) -> (b |$\Rightarrow$| tstar a).
Proof.
  apply tstar_ind; hauto lq:on inv:Par use:Par_refl,Par_cong ctrs:Par.
Qed.
\end{minted}
In prose, the triangle property can be proven by induction over the graph of
\mintinline{coq}{tstar a}, the Takahashi translation. Options \texttt{inv:Par}
and \texttt{ctrs:Par} say that the proof involves inverting and constructing
of the derivations of parallel reduction. The option
\texttt{use:Par\_refl,Par\_cong} allows the automation tactic to use the
reflexivity and congruence properties of parallel reduction as lemmas.

The flag \texttt{lq:on} tunes CoqHammer's search algorithm.  While this flag
appears arcane, when developing our proof scripts we never specify this option
manually. Instead, we first invoke the \texttt{best} tactic provided by
CoqHammer, specifying only the \texttt{inv}, \texttt{ctrs}, and lemmas that we
want to use. The \texttt{best} tactic then iterates through possible
configurations and provides us with a replacement with the tuned performance
flags that save time for future re-execution of the proof script.

The automation provided by CoqHammer not only gives us a proof that is shorter
and more resilient to changes, but also provides useful documentation for
readers who wish to understand the mechanized proof. Although automation
performs extensive search, we can configure it to not use lemmas or invert derivations that
are not specified in the \texttt{use} or \texttt{inv} flags.

@@

\section{Extensions}
In this section, we discuss how the normalization
proof of \lang can be extended to incorporate features including
natural numbers, dependent pairs, cumulative
universes with subtyping, and $\eta$ law for functions. All these
extensions have been implemented in our development. More advanced
features such as type-directed $\eta$ laws are out of the scope of
this paper, and we instead point the readers to existing work for
techniques that can be readily adapted to our minimal development.

\subsection{Natural Numbers}
\begin{figure}
  \drules[T]{$[[G |- a : A]]$}{Typing for Naturals}{Nat, Zero, Suc, Ind}
  \drules[I]{$[[Interp I i A S]]$}{Logical Relation for Naturals}{Nat}
  \drules[N]{$[[NatOk a]]$}{Natural values}{Zero, Neu, Suc}
\caption{Rules for natural numbers}
\label{fig:nat}
\end{figure}

@@

\Rref{I-Nat} gives us the semantic
interpretation of natural number; a term $[[a]]$ semantically inhabits
$[[Nat]]$ if $[[a]]$ reduces to some term $[[b]]$ such that $[[NatOk
b]]$ holds.
The $\vdash_\mathbb{N}$ judgment inductively characterizes the
normal form of (potentially open) terms of type $[[Nat]]$ and is
particularly useful in the \rref*{T-Ind} case of the fundamental
theorem. Let $[[rho |= G]]$, the induction hypothesis $[[G |= c :
Nat]]$ tells us that $[[c { rho } =>+ b ]]$ for $[[NatOk b]]$. The proof
then follows by induction over the derivation of $[[NatOk b]]$. When
$[[b]]$ is neutral, we conclude the proof with
\nameref{lemma:adequacy}. When $[[b]]$ is $[[zero]]$, we conclude the
proof with the outer induction hypothesis $[[G |= a : A {zero / x}]]$.
When $[[b]]$ is a successor, we conclude the proof with the outer
induction hypothesis $[[G ++ x : Nat ++ y : A |= b : A { succ x / x
}]]$.

@@

@@

    \hline
    \lang{} (this work) & $\mathbb{N}$ & Id, Bool & U & $\checkmark$ & 1 
    & Consistency and normalization \\
    % \hline
    $\lambda^\theta$ & 0 & Id, Nat & U & $\XSolidBrush$ & 1 & Consistency \\
    Core Nuprl & $\mathbb{N}$ & W-Types & E & $\checkmark$ & 2 & Consistency \\
    NBE-in-Coq & 1 & Nat & T & $\XSolidBrush$ & 2 & Correctness of NBE \\
    $\lambda^{\Pi U\mathbb{N}}$ & 1 & Nat, $\Sigma$ & T & $\checkmark$ & 2 & Decidability of conversion \\
    MLTT-\'a-la-Coq & 1 & Id, Nat, $\Sigma$ & T & $\checkmark$ & 2 & Decidibility of type checking \\
    \hline
  \end{tabular}

@@

\section{Conclusion}
\label{sec:conclusion}
In this work, we present a short and mechanized proof by logical relations for
a dependently typed language with a full universe hierarchy, large
eliminations, an intensional identity type, and dependent eliminators.  We
show the extensibility of our approach by proving the existence of
$\beta\eta$-normal forms with only small and mechanical changes to our proof
development.  Our Coq mechanization leverages existing Coq libraries for
reasoning about metatheory and for general purpose automation, allowing us to
significantly reduce the verbosity typically associated with mechanized
proofs. The result is a declarative proof style that rivals pen and paper.

@@

% Type soundness can be proven through a syntactic
% approach~\citep{syntacticsoundness} as a corollary of two properties:
% progress and preservation. % The syntactic type soundness proof
% % varies in complexity depending on the underlying type
% % system. For example, a type system that tracks information flow would
% % require additional structural rules related to security levels. In
% % this paper, we focus on one specific type of complexity: the
% In Figure~\ref{fig:stlcsoundness}, we summarize the structure of the
% syntactic type soundness proof for the simply typed lambda
% calculus. Each lemma can be proven by structural induction over the
% typing derivation, while using the previous established results as
% lemmas for specific cases that do not immediately follow from the
% induction hypothesis. If we make our language more complex by adding
% full dependent type support, the overall structure remains almost
% identical.

% NbE in Coq

>>>>

\jc{Delete this section}

@@

% The $\beta\eta$-normalization proof from Section~\ref{sec:betaeta} comprises
% 1549 lines of non-blank, non-comment lines of code. We choose not to include
% it in the chart, because of slight differences in lemma dependencies for
% untyped reduction and normal forms that make the comparison less
% informative. However, when compared to the $\beta$-normalization extension,
% the $\beta\eta$ extension has the same line count in the definition of the
% logical relation and the semantic soundness proof.

@@

We use the Autosubst 2 framework to produce Coq syntax files based on a de
Bruijn representation of variable binding and capture-avoiding substitution.
In addition to these generated definitions, Autosubst 2 provides a powerful
\texttt{asimpl} tactic used to prove the equivalence of two terms
constructed using the primitive operators provided by the framework. This
tactic simplifies the reasoning about substitution as many
substitution-related properties about syntax are immediately discharged by
\texttt{asimpl}.

For other automation tasks that are not specific to binding, we use
the powerful \texttt{sauto} tactic provided by CoqHammer to write
short and declarative proofs. For example, here is a one-line proof of
the triangle property about parallel reduction, from which the diamond
property (Lemma~\ref{lemma:pardiamond}) follows as a corollary.
The triangle property states
that if $[[a => b]]$, then $[[b]]\Rightarrow [[a]]^*$, where $[[a]]^*$
is the Takahashi translation~\citep{takahashi-parallel-reduction},
which roughly corresponds to simultaneous reduction of the redexes in
$[[a]]$, excluding the new redexes that appear as a result of
reduction.
\begin{minted}[escapeinside=||,mathescape=true]{coq}
Lemma Par_triangle a : forall b, (a |$\Rightarrow$| b) -> (b |$\Rightarrow$| tstar a).
Proof.
  apply tstar_ind; hauto lq:on inv:Par use:Par_refl,Par_cong ctrs:Par.
Qed.
\end{minted}
In prose, the triangle property can be proven by induction over the graph of
\mintinline{coq}{tstar a}, the Takahashi translation. Options \texttt{inv:Par}
and \texttt{ctrs:Par} say that the proof involves inverting and constructing
of the derivations of parallel reduction. The option
\texttt{use:Par\_refl,Par\_cong} allows the automation tactic to use the
reflexivity and congruence properties of parallel reduction as lemmas.

The flag \texttt{lq:on} tunes CoqHammer's search algorithm, which is never
specified manually. Instead, we first invoke the \texttt{best} tactic provided by
CoqHammer, specifying only the \texttt{inv}, \texttt{ctrs}, and lemmas that we
want to use. The \texttt{best} tactic then iterates through possible
configurations and provides us with a replacement with the tuned performance
flags that save time for future re-execution of the proof script.

The automation provided by CoqHammer not only gives us a proof that is shorter
and more resilient to changes, but also provides useful documentation for
readers who wish to understand the mechanized proof. Although automation
performs extensive search, we can configure it to only use lemmas or invert
derivations specified by the \texttt{use} or \texttt{inv} flags.

\jc{Just a digression but I find that CoqHammer automated proofs are \emph{not}
good documentation compared to how we might have otherwise written the proofs
manually. While the flags indicate what lemmas and inversions \emph{may} be used,
the \texttt{best} tactic does not delete unused ones, so there's no way of knowing
which ones actually \emph{are} used, and the automation hides away \emph{how} they
are used. Automation makes proofs resilient to change when things go right, but
when things go wrong, it becomes much harder to discern what lemmas I need to add
when I don't even know where the existing ones are used.}

@@

\section{Extensions}

In this section, we discuss how the normalization
proof of \lang can be extended to incorporate features including
natural numbers, dependent pairs, cumulative
universes with subtyping, and the $\eta$ law for functions. These
extensions have been implemented in our development. More advanced
features such as type-directed $\eta$ laws are out of the scope of
this paper, and we instead point the readers to existing work for
techniques that can be readily adapted to our minimal development.

\jc{We may want to briefly mention the LOC for each addition?}

\subsection{Natural Numbers}
\begin{figure}
  \drules[T]{$[[G |- a : A]]$}{Typing (naturals)}{Nat, Zero, Suc}
  \vspace{-\baselineskip}
  \begin{mathpar}
    \drule[width=\textwidth]{T-Ind}
  \end{mathpar}
  \drules[I]{$[[Interp I i A S]]$}{Logical relation (naturals)}{Nat}
  \drules[N]{$[[NatOk a]]$}{Natural values}{Zero, Neu, Suc}
\caption{Rules for natural numbers}
\label{fig:nat}
\end{figure}

@@

\Rref{I-Nat} gives us the semantic interpretation of natural number:
a term $[[a]]$ semantically inhabits $[[Nat]]$ if $[[a]]$ reduces to
some term $[[b]]$ such that $[[NatOk b]]$ holds.
The $[[NatOk b]]$ judgment inductively characterizes the
normal form of (potentially open) terms of type $[[Nat]]$ and is
particularly useful in the \rref*{T-Ind} case of the fundamental
theorem. Let $[[rho |= G]]$, the induction hypothesis $[[G |= c :
Nat]]$ tells us that $[[c { rho } =>+ b ]]$ for $[[NatOk b]]$. The proof
then follows by induction over the derivation of $[[NatOk b]]$. When
$[[b]]$ is neutral, we conclude the proof with
\nameref{lemma:adequacy}. When $[[b]]$ is $[[zero]]$, we conclude the
proof with the outer induction hypothesis $[[G |= a : A {zero / x}]]$.
When $[[b]]$ is a successor, we conclude the proof with the outer
induction hypothesis $[[G ++ x : Nat ++ y : A |= b : A { succ x / x
}]]$.

@@

\subsection{\texorpdfstring{$\beta\eta$}{beta-eta}-normal forms}

@@

    \hline
    \lang{} (this work) & $\mathbb{N}$ & Id, Bool & U & $\checkmark$ & 1 
    & Consistency and normalization \\
    % \hline
    $\lambda^\theta$ & 0 & Id, Nat & U & $\XSolidBrush$ & 1 & Consistency \\
    Core Nuprl & $\mathbb{N}$ & W-Types & E & $\checkmark$ & 2 & Consistency \\
    NBE-in-Coq & 1 & Nat & T & $\XSolidBrush$ & 2 & Correctness of NBE \\
    $\lambda^{\Pi U\mathbb{N}}$ & 1 & Nat, $\Sigma$ & T & $\checkmark$ & 2 & Decidability of conversion \\
    MLTT-\`a-la-Coq & 1 & Id, Nat, $\Sigma$ & T & $\checkmark$ & 2 & Decidibility of type checking \\
    \hline
  \end{tabular}

@@

\section{Conclusion}
\label{sec:conclusion}
In this work, we present a short and mechanized proof by logical relations for
a dependently typed language with a predicative universe hierarchy, large
elimination, an intensional identity type, and dependent eliminators.
From the fundamental theorem for the logical relation, we prove consistency
and canonicity, then show the extensibility of our approach by proving the
existence of $\beta$($\eta$)-normal forms and adding natural numbers and
cumulativity to the language. These extensions only require small and mechanical
changes to our proof development. Our Coq mechanization leverages existing Coq
libraries for reasoning about metatheory and for general purpose automation,
allowing us to significantly reduce the verbosity typically associated with
mechanized proofs. The result is a declarative proof style that rivals pen and paper.

@@
